\vspace{-1.5em} \bf Towards Pandemic-Scale Ancestral Recombination Graphs of SARS-CoV-2

### 2023-06-08 paper.tex
Updates from Overleaf

\section{Results}
\subsection{Inferred ARGs}
The goals of this preliminary study are to illustrate the utility of
\texttt{sc2ts} and to investigate the properties of the
inferred ARGs to inform subsequent development.
We work with a
representative subset of the available data, limited to
inferences that can be performed on a single server in
a few weeks (see below for further details on timings and
computer hardware used). The cut-off dates for sampling
are arbitrary.
We inferred two ARGs, which we refer to as the
``Wide'' and ``Long'' ARGs throughout.
The Wide ARG is densely sampled but time-limited and
includes 1.27 million strains collected up to June 30, 2021
which
pass some quality-control filters (Section~\ref{sec:data_preprocessing})
and  have a maximum delay between sampling and submission dates of 30 days
(Section~\ref{sec:filtering_time_travellers}).
For the Long ARG, we randomly sub-sample a maximum of 1,000 genomes per
day (again restricting the delay between sampling and submission to 30 days)
and include an additional year's worth of samples (to June 30, 2022).

@@

\subsection{False positive recombinants}
\label{sec:false_positives}
The Long and Wide ARGs both contain several recombination nodes that are
ancestral to a large number of samples.
These inferences of recombination may well be artefacts
caused by the appearance of new variants of concern carrying
more than the expected number of mutations~\citep{otto2021origins}.
Table~\ref{tab:false_positive} shows details of four recombinants
in the Long ARG that are likely false positives and have come to be
the ancestors of a large number of samples. These are the top-four recombinants
from the Long ARG in terms of numbers of descendant samples
(the fifth-largest has substantially fewer, at 14,869 descendant samples),
labelled A--D.
For each row, we show the sample's Pango lineage
designation and the rank of that sample in the overall list of
samples with that designation. Thus, rows C and D are the first
samples seen in the Long ARG from the BA.2.9 and BA.3 lineages,
and A and B are the third and eighth samples from the B.1.617.2
lineage (Delta VoC), respectively. Also shown is the overall ``cost'' of the Viterbi
solution computed by the LS HMM (Section~\ref{sec:ls}), which
is $3 \times$ number of recombinations + number of additional mutations
(for a mismatch ratio of $k=3$).
This column shows that each of these samples was a large
``distance'' from the current ARG when it was added. The mean
HMM cost over all 2,078 recombinants in the Long ARG is 8.39
(median: 7), and the mean cost for the  763 non-singleton
recombinants is 8.5 (median: 7). Thus, the recombinants in
Table~\ref{tab:false_positive} are in the tail of this distribution.

\begin{table} \centering
\begin{tabular}{llllll}
\toprule
& Strain & Descendants & Lineage & Rank in lineage&HMM Cost \\
\midrule
A&Germany/HH-RKI-I-061284/2021 & 178,405    & B.1.617.2   & 3& 17  \\
B&India/ILSGS00961/2021        & 177,649    & B.1.617.2   & 8& 25   \\
C&Denmark/DCGC-281594/2021     & 127,227    & BA.2.9      & 1& 32 \\
D&USA/NJ-GBW-EWR000001/2021    & 127,230    & BA.3        & 1& 25 \\
\bottomrule
\end{tabular}
\caption{\label{tab:false_positive}
False positive recombination events. Details show are for the top
four recombination nodes in the Long ARG ordered by number of
descending samples. See the text for details on the remaining columns.
Rows are labelled A, B, C and D for ease of reference.}
\end{table}

@@

\begin{figure} \centering
\includegraphics[width=0.5\textwidth]{figures/ls_diagram_covid.pdf}
\caption{\label{fig:ls_diagram} A schematic of the Li and Stephens (LS)
model, in which a focal sequence (bottom) is described as an
imperfect mosaic of the sequences in a reference panel.
Black crosses along the focal sequence show sequencing
errors or mutations.
In the standard formulation, at site $\ell$, the recombination probability is $r_\ell$,
the mutation probability is $\mu_\ell$ and $n$
denotes the size of the reference panel.
The Viterbi algorithm can be used to find a
``copying path'' through the reference panel for a given focal sequence that
maximises the likelihood under these parameters. Unseen states in the reference panel are shown as coloured lines enclosed by
the grey box. The black arrow describes the true path through the data which leads to the emitted
focal sequence below. Examples of transition and
emission probabilities along this trajectory are shown by the red and blue
arrows, respectively.
}
\end{figure}

@@

\begin{figure} \centering
\includegraphics[width=0.85\textwidth]{figures/false_positive_top2_nxcld_large_graph.pdf}
\caption{\label{fig:false_positives}
Subgraph of the Long ARG, focussing on two likely false positive
recombination nodes at the start of the Delta wave
(``focal'' nodes, in red, corresponding to rows A and B in
Table~\ref{tab:false_positive}).
The path to node \texttt{tsk261771} has been expanded:
this node (in gold) is the ancestor of ${\sim}89.8\%$ of Delta samples in the Long ARG,
and represents a large polytomy with 107 immediate children.
The suspected-incorrect recombination path to an early B.1 MRCA is also shown.
Note the large amount of additional recombination (black nodes) among close
descendants of the focal nodes.
% Why say this here and not at for the other subgraph figs?
%See supplementary file \protect\path{Subgraph_sample_mapping.txt} to  map tskID to EPI\_ISL and strain name.
}
\end{figure}

>>>>

\section{Results}
\subsection{Inferred ARGs}
The goals of this preliminary study are to illustrate the utility of
\texttt{sc2ts} and to investigate the properties of the
inferred ARGs to inform subsequent development.
We work with a
representative subset of the available data, limited to
inferences that can be performed on a single server in
a few weeks (see below for further details on timings and
computer hardware used). The cut-off dates for sampling
are arbitrary.
We inferred two ARGs, which we refer to as the
``Wide'' and ``Long'' ARGs throughout.
The Wide ARG is densely sampled but time-limited and
includes 1.27 million sequences collected up to June 30, 2021
which
pass some quality-control filters (Section~\ref{sec:data_preprocessing})
and  have a maximum delay between sampling and submission dates of 30 days
(Section~\ref{sec:filtering_time_travellers}).
For the Long ARG, we randomly sub-sample a maximum of 1,000 genomes per
day (again restricting the delay between sampling and submission to 30 days)
and include an additional year's worth of samples (to June 30, 2022).

@@

\subsection{False positive recombinants}
\label{sec:false_positives}
The Long and Wide ARGs both contain several recombination nodes that are
ancestral to a large number of samples.
These inferences of recombination may well be artefacts
caused by the appearance of new variants of concern carrying
more than the expected number of mutations~\citep{otto2021origins}.
Table~\ref{tab:false_positive} shows details of four recombinants
in the Long ARG that are likely false positives and have come to be
the ancestors of a large number of samples. These are the top-four recombinants
from the Long ARG in terms of numbers of descendant samples
(the fifth-largest has substantially fewer, at 14,869 descendant samples),
labelled A--D.
For each row, we show the sample's Pango lineage
designation and the temporal rank of that sample out of all
samples with that designation. Thus, rows C and D are the earliest
samples seen in the Long ARG from the BA.2.9 and BA.3 lineages,
and A and B are respectively the third and eighth earliest samples among 12829 samples
assigned by Nextclade to the B.1.617.2
lineage (Delta VoC). Also shown is the overall ``cost'' of the Viterbi
solution computed by the LS HMM (Section~\ref{sec:ls}), which
is $3 \times$ number of recombinations + number of additional mutations
(for a mismatch ratio of $k=3$).
This column shows that each of these samples was a large
``distance'' from the current ARG when it was added. The mean
HMM cost over all 2,078 recombinants in the Long ARG is 8.39
(median: 7), and the mean cost for the  763 non-singleton
recombinants is 8.5 (median: 7). Thus, the recombinants in
Table~\ref{tab:false_positive} are in the tail of this distribution.

\begin{table} \centering
\begin{tabular}{llllll}
\toprule
& Strain & Descendants & Lineage & Rank in lineage&HMM Cost \\
\midrule
A&Germany/HH-RKI-I-061284/2021 & 178,405    & B.1.617.2   & 3 / 12829& 17  \\
B&India/ILSGS00961/2021        & 177,649    & B.1.617.2   & 8 / 12829& 25   \\
C&Denmark/DCGC-281594/2021     & 127,227    & BA.2.9      & 1 / 10897& 32 \\
D&USA/NJ-GBW-EWR000001/2021    & 127,230    & BA.3        & 1 / 15& 25 \\
\bottomrule
\end{tabular}
\caption{\label{tab:false_positive}
False positive recombination events. Details show are for the top
four recombination nodes in the Long ARG ordered by number of
descending samples. See the text for details on the remaining columns.
Rows are labelled A, B, C and D for ease of reference.}
\end{table}

@@

\begin{figure} \centering
\includegraphics[width=0.5\textwidth]{figures/Li_S.pdf}
\caption{\label{fig:ls_diagram} A schematic of the Li and Stephens (LS)
model, in which a focal sequence (bottom) is described as an
imperfect mosaic of the sequences in a reference panel.
Black crosses along the focal sequence show sequencing
errors or mutations.
In the standard formulation, at site $\ell$, the recombination probability is $r_\ell$,
the mutation probability is $\mu_\ell$ and $n$
denotes the size of the reference panel.
The Viterbi algorithm can be used to find a
``copying path'' through the reference panel for a given focal sequence that
maximises the likelihood under these parameters. Unseen states in the reference panel are shown as coloured lines enclosed by
the grey box. The black arrow describes the true path through the data which leads to the emitted
focal sequence below. Examples of transition and
emission probabilities along this trajectory are shown by the red and blue
arrows, respectively.
}
\end{figure}

@@

\begin{figure} \centering
\includegraphics[width=0.85\textwidth]{figures/false_positive_top2_nxcld_large_graph.pdf}
\caption{\label{fig:false_positives}
Subgraph of the Long ARG, focussing on two likely false positive
recombination nodes at the start of the Delta wave
(``focal'' nodes, in red, corresponding to rows A and B in
Table~\ref{tab:false_positive}).
The path to node \texttt{tsk261771} has been expanded:
this node (in gold) is the ancestor of ${\sim}89.8\%$ of Delta samples in the Long ARG,
and represents a large polytomy with 107 immediate children. Of the remaining Delta samples, 10\% of Delta
samples are mostly descendants of the node \texttt{tsk232088} on the far right,
a sibling of the earliest focal recombination node.
The suspected-incorrect recombination path to an early B.1 MRCA is also shown.
Note the large amount of additional recombination (black nodes) among close
descendants of the focal nodes.
% Why say this here and not at for the other subgraph figs?
%See supplementary file \protect\path{Subgraph_sample_mapping.txt} to  map tskID to EPI\_ISL and strain name.
}
\end{figure}
